---
title: "Assignment 3 - CRD"
author: "UMAR MUNIR"  
format:
  html:
    code-fold: false
    embed-resources: true
    toc: true
    number-sections: false
    theme: cerulean
    
---

# Introduction  
This assignment will ask you to conduct a full CRD analysis, from data import through publication ready plot.  

For that, you will first need to watch (and code along with) the following CRD videos:

- CRD lecture: https://youtu.be/r5vsakgqMXc?si=8udshzlpZ-8td1f4&t=4941 (starts at 1:22:21, watch till end)

- CRD lecture pt 2, lab pt 1: https://youtu.be/GpfHxh1eK5U?si=nTle5YcB8JQklja2 
(watch whole video)  

- CRD lab pt 2: https://youtu.be/YQ1bK6uV194?si=LnDzRT69OsiMLc-r
(wathc the whole video)  

**AFTER watching the videos and developing the code in class**, then start working on this HW assignment.  

The different sections representing what we did in class were left in the script to provide you a scaffold to start from.  

# Instructions  
  - Move this script into the `code` subfolder of your `03_crd` project folder  
  - Move the file `wheat_nk_bamyan.csv` into the `data` folder of your `03_crd` project folder  
  - On this script, fill your name under the `title` section of the YAML  

# a) Setup  
Here is where we load the packages we will use.  
Make sure to load each package in the chunk below.  
```{r setup}

#install.packages("broom")
#install.packages("multcompView")
#Loading packages
library(tidyverse) # data wranggling and plotting
library(car) # Anova functions
library(emmeans) # model residuals extraction
library(multcomp) # pairwise comparison letter display
library(broom)

```

Now, import the data set `wheat_nk_bamyan.csv` in the chunk below.  
```{r data import}
crd_df <- read_csv("../data/wheat_nk_bamyan.csv")

crd_df
#check how R is interpreting everything as numeric. We need to change some columns types.

```

The **objective of this study** was to assess the effect of different N and K rates on wheat yield.  

For that, a study was conducted using a completely randomized design as the experimental design with four replicates. The treatment design was a 3 N x 3 K rate factorial, for a total of 9 treatments.  


# b) EDA tables  
Use this space to conduct exploratory data analysis of the imported data, in the form of tables.

```{r summary}
  summary(crd_df)

```


```{r glimpse}
glimpse(crd_df)
```


# c) Wrangling  
Here, make sure to wrangle your data set with all needed steps. Refer back to class code if you need a refresher on what was needed.  

```{r wrangling}

crd_dfw <- crd_df %>% 
  mutate(rep = factor(rep),
         nrate_kgha = factor(nrate_kgha),
         krate_kgha = factor(krate_kgha)
         ) %>% 
  mutate(trtname = paste0(nrate_kgha, "+", krate_kgha))

crd_dfw

```

```{r wrangled data}
summary(crd_dfw)

```
 

# d) EDA plots  

## Question 1) Below each plot of this section, write a sentence explaining your visual interpretation of the plot.

Now, explore the wrangled data using plots.  
Make sure to explore plots showing the main effects and the interaction.  

```{r n boxplot}
ggplot(crd_dfw, aes(x = nrate_kgha,
                    y = yield_kgha,
                    color = nrate_kgha)) +
  geom_boxplot() +
  geom_jitter() +
  theme(legend.position = "none")


```
The boxplot shows an increasing yield trend with rising nitrogen rates, with a greater yield increase from 0-100 kg/ha than from 100-200 kg/ha, while the interquartile range remains similar.





```{r k boxplot}
ggplot(crd_dfw, aes(x = krate_kgha,
                    y = yield_kgha,
                    color = krate_kgha)) +
  geom_boxplot() +
  geom_jitter() +
  theme(legend.position = "none")
```
Based on visual interpretation, yield shows a positive relationship with increasing potassium rate, while the interquartile range varies among the boxplots, with 30 kg/ha having the greatest spread.

```{r nk boxplot}

ggplot(crd_dfw, aes(x = nrate_kgha,
                    y = yield_kgha,
                    color = nrate_kgha)) +
  geom_boxplot() +
  geom_jitter() +
  facet_grid(.~krate_kgha) +
  theme(legend.position = "none")
```
The individual positive effects of nitrogen and potassium are evident here as well, but their combined effect is significantly higher, with similar yields observed for 30 kg/ha potassium + 200 kg/ha nitrogen and 60 kg/ha nitrogen + 100 kg/ha potassium, while yield plateaus beyond the latter.


# e) Statistical model  
In this section, run the appropriate model.  
Remember to set the contrast type to sum-to-zero (the default in R is set-to-zero, which is not what we want).  

Also, remember to specify all model terms needed for the treatment design.  

```{r model}
#changing to sum-to-zero contrast
options (contrasts = c("contr.sum", "contr.poly"))

#Model_fitting
crd_mod <- lm(yield_kgha ~ nrate_kgha*krate_kgha,
              data = crd_dfw)

summary(crd_mod)





```



# f) ANOVA table  
Use this section to check the ANOVA table, ensuring you use type 3 sum-of-squares.


```{r anova}
Anova(crd_mod, type = 3)
```
## Question 2) Based on the ANOVA table above, for which effect(s) in the ANOVA should you extract means and perform pair-wise comparisons? Explain why.  
We should extract means for the main treatments i.e., individual nitrogen and potassium rates becuase they have significant effects. We should also perform pair-wise comparison for these treatmetns and not the interaction.

# g) Linear model assumptions  
Use this section to first compute your residuals, including studentdized residuals.  


```{r crd_resid}

crd_resid <- augment(crd_mod) %>% 
  mutate(.stdresid = rstudent(crd_mod))

crd_resid


```





## Question 3) What are the linear model assumptions that we need to check for? Create subsections (using ##) below for each of these assumptions, and write code to produce plots to check for these assumptions. For each assumption, write down your interpretation of the produced plots, and your thoughts if the assumptions hold or not.

The linear model assumptions that we need to check for are following,
1) Independence of the residuals (no pattern)
2) Presence of Homoscedacity
3) Normality of the residuals
4) No residual outliers

## Residual independence

```{r resid independence}

ggplot(crd_resid, aes(x = .fitted,
                      y = .stdresid)) +

  geom_point(shape = 21,
             fill = "purple",
             size = 3,
             alpha = .7) +
  geom_hline(yintercept = c(-3,0,3), color = "red") +
  geom_smooth() +
  theme_bw()

```
Interpretation: The residues are independent because the spread is uniform around the zero depicted by the grey shape. The blue line that is overlapping the red line at zero is depicting that there is no pattern in residuals hence independece is established.

##Homoscedacity

```{r homoscedacity}
ggplot(crd_resid, aes(x = .fitted,
                      y = .stdresid)) +

  geom_point(shape = 21,
             fill = "purple",
             size = 3,
             alpha = .7) +
  geom_hline(yintercept = c(-3,0,3), color = "red") +
  geom_smooth() +
  theme_bw()

```
Since the error bands are not showing fan like pattern or the error are not increasing, we can conclude that the variance is homogeneous across treatments. Therefore, homoscedacity is established.

## Residual noramlity

```{r resid normality}

ggplot(crd_resid, aes(sample = .stdresid))+
  stat_qq() +
  stat_qq_line() + theme_bw()
#point should be very close almost on top of the line. Some departure at tales is fine especialy data is small
```
The requirement for residuals to follow normality is that the residuals points should be close to the regression which is clearly the case, hence normality of residuals is established.




```{r resid normality}

ggplot(crd_resid, aes(x = .stdresid)) +
  geom_density() +
  scale_x_continuous(breaks = c(-3,0,3), limits = c(-3,3)) + theme_bw()

```
Although the bell curve is wider at the bottom, the majority of the data is centered around zero and within 3 standard deviation therefore we can conclude that normality is present in residuals.

## Residual outliers

```{r resid outliers}

ggplot(crd_resid, aes(x = .fitted,
                      y = .stdresid)) +

  geom_point(shape = 21,
             fill = "purple",
             size = 3,
             alpha = .7) +
  geom_hline(yintercept = c(-3,0,3), color = "red") +
  geom_smooth() +
  theme_bw()


```
There are no outliers in the data shown above. All the data points are within the 3 standard deviation lines (red lines) at top and bottom.
  

# h) Model means  
Use this section to extract model means according to your decision of the ANOVA table effects.  
extract means of significant trts

```{r interaction mean of nrate_kgha}

crd_means_n <- emmeans(crd_mod, 
                         ~nrate_kgha)

crd_means_n

```

```{r interaction mean of krate_kgha}

crd_means_k <- emmeans(crd_mod, ~ krate_kgha)

crd_means_k

```




# i) Pairwise comparisons  
Use this section to perform pair-wise comparison of the means extracted above.  

Make sure to reverse the order from high to low, use letters instead of numbers, and no adjustment.  

```{r interaction pwc all}

crd_cld_n <- cld(crd_means_n,
                   reversed = T,
                   Letters = letters,
                   adjust = "none")

crd_cld_n

```
```{r interaction n inside k}
crd_cld_k <- cld(crd_means_k,
                   reversed = T,
                   Letters = letters,
                   adjust = "none")

crd_cld_k

```




```{r selected pwc}

crd_cld_selected_n <- crd_cld_n %>% 
  as.data.frame() %>% 
  mutate( letter = trimws(.group))


crd_cld_selected_n
```

```{r}
crd_cld_selected_k <- crd_cld_k %>% 
  as.data.frame() %>% 
  mutate( letter = trimws(.group))


crd_cld_selected_k
```


# g) Final plot(s)  
Produce final plot(s) to show your results.  
Make sure to include both **raw data** and **statistical model summary (i.e., letter separation)**.  


```{r crd final plot}
#| fig-width: 8
#| fig-height: 5
ggplot(mapping = aes(fill = nrate_kgha)
       ) +

  geom_boxplot(data = crd_dfw,
               aes(x = nrate_kgha, 
                   y = yield_kgha),
                   alpha = .85) +
  geom_jitter(data = crd_dfw,
              aes(x = nrate_kgha, 
                   y = yield_kgha),
              shape = 21,
              alpha = .7,
              size = 3)+
  geom_label(data = crd_cld_selected_n, aes(x = nrate_kgha, 
                                          y = emmean,
                                          label = letter),
             fill = "white"
             ) +
  theme_classic() +
  labs(title = "Effect of nitrogen on wheat yield",
       x = "Nitrogen rate (kg/ha)",
       y = "Yield (kg/ha)") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(legend.position = "none")

```




```{r}
#| fig-width: 8
#| fig-height: 5
ggplot(mapping = aes(fill = krate_kgha)) +

  geom_boxplot(data = crd_dfw,
               aes(x = krate_kgha, 
                   y = yield_kgha),
               alpha = .85) +
  geom_jitter(data = crd_dfw,
              aes(x = krate_kgha, 
                   y = yield_kgha),
              shape = 21,
              alpha = .7,
              size = 3)+
  geom_label(data = crd_cld_selected_k, aes(x = krate_kgha, 
                                          y = emmean,
                                          label = letter),
             fill ="white") +
  theme_classic() +
  labs(title = "Effect of potassium on wheat yield",
       x = "Potassium rate (kg/ha)",
       y = "Yield (kg/ha)") +
  theme(plot.title = element_text(hjust = 0.5))+
    theme(legend.position = "none")


```






Be mindful of your plot to ensure you are creating it in an intuitive way for your audience in relation to interpreting the letter separation structure you used.  

In class, we did not have time to make it publication-ready (e.g., a plot looking professional and ready to be included in a manuscript).

Here, make sure you deliver final plot(s) that is/are publication ready.  

Make sure to provide an interpretation for the plot(s) created, including which treatment levels were significantly higher/lower.


# Submitting your work  
Once you have developed all the code and answers, make sure to Render this quarto file.  

**Notes on rendering**:  

- Make sure to render your work and inspect how the final html look like.  
- If it does not look professional for whatever reason, then fix the issue, re-render it, recheck.  
- Only send me your work once your html file looks professional.  
  - **DO NOT** delete the file's heading levels (# and ##). They set up the proper heading 1 and 2 levels, and I use them to guide my grading.  
  - If a given chunk is also outputting warnings or messages, inhibit this behavior by changing the chunk options `message` and `warning` to `FALSE` (I'll demo this in a moment).  
  
  - If, after rendered, 2 lines of text are connected and you wish to "break line" between them, add 2 extra spaces after the first one (I'lld demo this in a moment).  

After rendering, an .html file will be created on your `code` folder.  

Rename this file to `Assignment-03-crd-LASTNAME.html`.    

For ex., mine would be `Assignment-03-crd-Bastos.html`.

Submit your work by uploading the **html** file to **eLC Assigment #3 - CRD** by Feb 3rd 11:59 pm.  
  




